Radiance - A Web Application Environment

### 2017-03-17 els2017-radiance/paper.tex
Incorporated feedback from Robert Strandh

Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone. However, this can lead to issues when multiple applications should be deployed side-by-side. For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together. Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation, but instead run the risk of introducing too many features that remain unused. \\

Radiance initially grew out of the desire to write web applications that could be run together in such a way, that users would not have to create multiple accounts, track multiple logins, and so forth. Another primary goal of it was to write it in such a way, that it would be possible to exchange various parts of the framework without needing to change application code. This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.

@@

When talking about such an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver. This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application. \\

@@

\subsection{Interfaces}
Interfaces represent a form of contract. They allow you to define the signatures of constructs exposed from a package, such as functions, macros, classes, and so forth. These function signatures are accompanied by documentation that specifies their public behaviour. Through this, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them. \\

In order to make this more concrete, let's look at an interface that might provide some sort of caching mechanism.

@@

However, as it currently is, this will obviously not work. The \code{with-caching} macro is not implemented with any useful functionality. As the writer of the blog application, we don't need to know the specifics of how the caching is implemented, though. All we need to do is tell Radiance that we need this interface. This can be done by adding the interface as a dependency to our application's ASDF\cite{asdf} system definition.

@@

This is a particularly primitive and straight-forward implementation, in order to keep things short. The function and macro are provided by just overriding the stub definitions that were in place previously. The variable definition is not part of the official interface, and is instead exposed through an unexported symbol, denoting an implementation-dependant extension. \\

Using direct overwriting of definitions means that all applications must use the same implementation. However, usually this is the intended effect, as we want to maximise the sharing between them. If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on. This approach with interfaces does bring some benefits, however. For one, it allows the implementations to be as efficient as possible, as there is no intermediate layer of redirection. \\

@@

Finally, Radiance provides a system that allows to optionally depend on an interface. This is useful to model something like the user preferences mentioned in our example application. An administrator might not always want to provide an administration or settings panel. To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded. For an imaginary \code{user-settings} interface, this might look like the following:

@@

\subsection{Routes}
In order to allow the administrator to change where pages are accessible from, an application cannot hard-code its resource locations and the links in its templates. This is doubly important when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another. \\

In many frameworks, like for example Symfony\cite{symfony}, this is solved by naming every resource in the system by a tag, and then allowing the configuration of what each tag resolves to individually. Radiance takes a different approach. It introduces the idea of two separate namespaces: an external one, which is what a visitor of a website sees and interacts with, and an internal one, which is what the programmer of an application deals with. \\

@@

When a request is dispatched by Radiance, it first parses the request URL into an object presentation that makes it easier to modify. It then sends it through the routing system's mapping functions, which turn this external URL into an internal one. The request is then dispatched to the application that ``owns'' the URL in the internal representation. \\

@@

If, as an administrator of an installation of this blog application, we now wanted to reach this page through the URL \texttt{www.example.com/blog/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{blog} folder, and put that as the URL's domain instead. We can achieve this relatively easily with the following definition.

@@

The templating engine takes care of reading out the \texttt{@} attribute values, running them through the routing system's reversal functions, and putting the resulting URL into the corresponding attribute to produce a valid link.

@@

We need to stick this into a trigger, in order to defer the evaluation to when Radiance is being started up and the environment has been decided. During the loading of the system, the environment might not have been set yet, and we would not be able to access the configuration storage. \\

While Radiance does provide default implementations for all of its interfaces, it is likely that some of them are not usable for a production setting. In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file. We can either modify the file directly like so:

@@

Every application's individual configuration can be changed in much the same way.

@@

Through the environment system, Radiance standardises the way each application is configured by the administrator and how the system is pieced together when it is loaded. As a consequence of that it becomes trivial to switch between different setups of a Radiance installation. The simple, human-readable configuration format used allows even users without intimate knowledge of Lisp to set up an instance.

\section{Further Work}
Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them. Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process. However, if a constraint on what kind of changes can be rolled back is introduced, it should be possible to provide a usable form of this kind of feature. \\

Radiance also does not allow you to change the implementation of an interface on the fly. This has much the same problems as the previous issue. In addition, though, the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect. Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application. \\

@@

\section{Acknowledgements}
Thanks to Joram Schrijver, Till Ehrengruber, and Fernando Borretti for feedback and suggestions on the paper.

>>>>

Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone. However, this paradigm can lead to issues when multiple applications should be deployed side-by-side. For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together. Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation, but instead run the risk of introducing too many features that remain unused. \\

Radiance initially grew out of the desire to write web applications that could be run together in such a way, that users would not have to create multiple accounts, track multiple logins, and so forth. Another primary goal of it was to write it in such a way, that it would be possible to exchange various parts of the framework without needing to change application code. This separation would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.

@@

When talking about such an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver. This approach is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application. \\

@@

\subsection{Interfaces}
Interfaces represent a form of contract. They allow you to define the signatures of constructs exposed from a package, such as functions, macros, classes, and so forth. These function signatures are accompanied by documentation that specifies their public behaviour. Through this system, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them. \\

In order to make this system more concrete, let's look at an interface that might provide some sort of caching mechanism.

@@

However, as it currently is, this function will obviously not work. The \code{with-caching} macro is not implemented with any useful functionality. As the writer of the blog application, we don't need to know the specifics of how the caching is implemented, though. All we need to do is tell Radiance that we need this interface. This can be done by adding the interface as a dependency to our application's ASDF\cite{asdf} system definition.

@@

This implementation is a particularly primitive and straight-forward one, in order to keep things short. The function and macro are provided by just overriding the stub definitions that were in place previously. The variable definition is not part of the official interface, and is instead exposed through an unexported symbol, denoting an implementation-dependant extension. \\

Using direct overwriting of definitions means that all applications must use the same implementation. However, usually this effect is intended, as we want to maximise the sharing between them. If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on. This approach with interfaces does bring some benefits, however. For one, it allows the implementations to be as efficient as possible, as there is no intermediate layer of redirection. \\

@@

Finally, Radiance provides a system that allows the programmer to optionally depend on an interface. This mechanism is useful to model something like the user preferences mentioned in our example application. An administrator might not always want to provide an administration or settings panel. To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded. For an imaginary \code{user-settings} interface, an example might look like the following:

@@

\subsection{Routes}
In order to allow the administrator to change where pages are accessible from, an application cannot hard-code its resource locations and the links in its templates. It is doubly important to do this when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another. \\

In many frameworks, like for example Symfony\cite{symfony}, this problem is solved by naming every resource in the system by a tag, and then allowing the configuration of what each tag resolves to individually. Radiance takes a different approach. It introduces the idea of two separate namespaces: an external one, which is what a visitor of a website sees and interacts with, and an internal one, which is what the programmer of an application deals with. \\

@@

When a request is dispatched by Radiance, it first parses the request URL into an object presentation that makes it easier to modify. It then sends it through the mapping function of the routing system, which turn this external URL into an internal one. The request is then dispatched to the application that ``owns'' the URL in the internal representation. \\

@@

If, as an administrator of an installation of this blog application, we now wanted to reach this page through the URL \texttt{www.example.com/blog/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{blog} folder, and put that as the URL's domain instead. We can achieve this behaviour relatively easily with the following definition.

@@

The templating engine takes care of reading out the \texttt{@} attribute values, running them through the reversal functions of the routing system, and putting the resulting URL into the corresponding attribute to produce a valid link.

@@

We need to stick the configuration update into a trigger, in order to defer the evaluation to when Radiance is being started up and the environment has been decided. During the loading of the system, the environment might not have been set yet, and we would not be able to access the configuration storage. \\

While Radiance does provide default implementations for all of its interfaces, it is likely that some of them are not usable for a production setting. In order to change, say, the implementation of the database interface, we would then have to modify the Radiance core's configuration file. We can either modify the file directly like so:

@@

The individual configuration of every application can be changed in much the same way.

@@

Through the environment system, Radiance standardises the way each application is configured by the administrator and how the system is pieced together when it is loaded. As a consequence of that standardisation it becomes trivial to switch between different setups of a Radiance installation. The simple, human-readable configuration format used allows even users without intimate knowledge of Lisp to set up an instance.

\section{Further Work}
Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them. Doing so is troublesome, as the system of an application might potentially modify any other part of the Lisp process. However, if a constraint on what kind of changes can be rolled back is introduced, it should be possible to provide a usable form of this kind of feature. \\

Radiance also does not allow you to change the implementation of an interface on the fly. This feature has much the same problems as the previous issue. In addition, though, the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect. Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application. \\

@@

\section{Acknowledgements}
Thanks to Joram Schrijver, Till Ehrengruber, and Robert Strandh for feedback, corrections, and suggestions on the paper.
